---
title: "Cyclistic_Clean_Data"
author: "Trang Nguyen"
date: '2022-04-26'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Packages
```{r loading packages}
library(tidyverse)
library(dplyr)
library(readr)
library(janitor)
```

## Concatenating 
All the csvs file will be concatenated into one dataframe
```{r concatenating all csvs file into one dataframe}
csv_files <- list.files(path = "/Users/trangnguyen/Desktop/Cyclistics", 
                        recursive = TRUE, full.names=TRUE)
cyclistic_merged <- do.call(rbind, lapply(csv_files, read.csv))
head(cyclistic_merged)
```

## Remove duplicates
```{r}
cyclistic_no_dups <- cyclistic_merged[!duplicated(cyclistic_merged$ride_id),]
print(paste("Removed", nrow(cyclistic_merged)-nrow(cyclistic_no_dups), 
            "duplicated rows"))
```

## Parsing datetime columns
```{r}
cyclistic_no_dups$started_at <- as.POSIXct(cyclistic_no_dups$started_at, 
                                           "%Y-%m-%d %H:%M:%S")
cyclistic_no_dups$ended_at <- as.POSIXct(cyclistic_no_dups$ended_at, 
                                         "%Y-%m-%d %H:%M:%S")
```

## Manipulating the data
Adding new columns to help improve calculation time 

## ride_time_m 
Represents the total time of a bike ride, in minutes

```{r}
cyclistic_no_dups <- cyclistic_no_dups %>% 
  mutate(ride_time_m = as.numeric(cyclistic_no_dups$ended_at 
                                  - cyclistic_no_dups$started_at) / 60)
summary(cyclistic_no_dups$ride_time_m)
```

## year_month 
Separate the year and the month into one column

```{r}
cyclistic_no_dups <- cyclistic_no_dups %>% 
  mutate(year_month = paste(strftime(cyclistic_no_dups$started_at, "%Y"),
                            "-",
                            strftime(cyclistic_no_dups$started_at, "%m"),
                            paste("(",strftime(cyclistic_no_dups$started_at, 
                                               "%b"), ")", sep=" ")))
unique(cyclistic_no_dups$year_month)
```

## weekday
Determine patterns of travels in the week

```{r}
cyclistic_no_dups <- cyclistic_no_dups %>%
  mutate(weekday = paste(strftime(cyclistic_no_dups$ended_at, "%u"), "-",
                          strftime(cyclistic_no_dups$ended_at, "%a")))
unique(cyclistic_no_dups$weekday)
```

## start_hour
Getting the hour of the day also may be useful for intra day analysis

```{r}
cyclistic_no_dups <- cyclistic_no_dups %>% 
  mutate(start_hour = strftime(cyclistic_no_dups$ended_at, "%H"))
unique(cyclistic_no_dups$start_hour)

```

## Saving the result as a CSV
```{r}
cyclistic_no_dups %>% 
  write.csv("cyclistic_clean.csv")
```

## Resize the plot for analysis
```{r}
fig <- function(width,height){options(repr.plot.width, repr.plot.height = height)}
```
```{r}
cyclistic <- cyclistic_no_dups
head(cyclistic)
```

## Summary of the dataset
```{r}
summary(cyclistic)
```

