---
title: "Cyclistic_analyze_data"
author: "Trang Nguyen"
date: '2022-04-26'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading packages for analysis
I will be using ggplot2 for data visualization and analysis

```{r loading packages}
library(tidyverse)
library(ggplot2)
```

## Load the dataset

```{r}
cyclistic <- read_csv("cyclistic_clean.csv")
```


## Casuals vs Members Distribution

```{r casuals vs members distribution}
cyclistic %>% 
  group_by(member_casual) %>% 
  summarise(count = length(ride_id), 
            '%' = (length(ride_id) / nrow(cyclistic)) * 100 )
```
## Plot 1: Casuals x Members Distribution  
This visualization shows Casuals vs Members distribution which Casuals are ~45% and Members are ~55%. Casuals percentage is 10% lesser than the count of Members. 

```{r plot 1}
ggplot(cyclistic, aes(member_casual, fill = member_casual)) + geom_bar() + 
  labs(x = "Casuals x Members", title = "Plot 1: Casuals x Members Distribution") +
  scale_fill_brewer(palette = "Set2")
```
## Month Distribution
I will remove the 2021 - 03 ( Mar ) since this infomation is not valid/complete.

```{r month distribution}
cyclistic %>% 
  group_by(year_month) %>% 
  summarise(count = length(ride_id),
            '%' = length(ride_id) / nrow(cyclistic) * 100,
            'member_percent' = (sum(member_casual == "member") / length(ride_id)) *100,
            'casual_percent' = (sum(member_casual == "casual") / length(ride_id))* 100,
            'Member x Casual Perc Difer' = member_percent - casual_percent) 
```

## Remove incomlete/valid data
Removed "2021 - 03 ( Mar )" using the SUBSET fucntion since this information 
is not complete and/or invalid for analysis

```{r remove incomplete data}
cyclistic_new <- cyclistic %>% 
  subset(year_month != "2021 - 03 ( Mar )",)
```

## Plot 2.1: Month Distribution
This visualization shows how Casual vs Member distribute throughout the month
in the previous 12 months. Some key points taken from this plot:
1. The months with the biggest volume of rides are July and August.
2. The months with the least volume of rides are January and February . 
3. In all months, we tend to have more members' rides than casual rides. 
4. Our volume of rides is high during the summer months (casuals' volume of
rides also high during these few months) 
5. Let's compare this data with the climate data of Seattle to see whether we 
can identify any relationship. 

```{r plot 2.1}
cyclistic_new %>% 
  ggplot(aes(year_month, fill = member_casual)) + geom_bar () + 
  labs(x = "Month", title="Plot 2.1: Month Distribution") + coord_flip() + 
  scale_fill_brewer(palette = "Set2")
```

## Climate data in Seattle 
I will compare the above data with the climate data from Seattle to see if I can 
draw further details from the Casual vs Member Month Distribution.
The data is from <https://en.wikipedia.org/wiki/Climate_of_Seattle> 

```{r seattle_temp}
seattle_mean_temp <- c(6.0, 6.7, 8.4, 10.7, 14.2, 16.7, 19.5, 19.7, 17.0, 12.1, 8.1, 5.6)
month <- c("01 - Jan","02 - Feb","03 - Mar","04 - Apr","05 - May","06 - Jun","07 - Jul","08 - Aug","09 - Sep","10 - Oct","11 - Nov","12 - Dec")
```

## Plot 2.2: Seattle Temperature
Based on the Seattle climate data, there's a significant relationship between the temperature and the volume of rides throughout the year. 


```{r plot 2.2}
data.frame(month, seattle_mean_temp) %>% 
  ggplot(aes(x = month, y = seattle_mean_temp, fill = month)) +
  labs(x= "Month", y = "Mean temperature", title = "Plot 2.2: Mean Temperature for Seattle (1991-2020)") + geom_col() 
```

## Weekday Distribution

```{r weekday distribution}
cyclistic_new %>% 
  group_by(weekday) %>% 
  summarise(count = length(ride_id),
            '%' = (length(ride_id) / nrow(cyclistic_new)) * 100,
            'members_percent' = (sum(member_casual == "member") /
                                   length(ride_id)) *100,
            'casual_percent' = (sum(member_casual == "casual") / 
                                  length(ride_id)) * 100,
            'Member x Casual Perc Difer' = members_percent - casual_percent)
```

## Plot 3: Weekday Distribution
Based on the data, there's a few key points:
1. The biggest volume of rides during the week is on the weekend with Saturday's 
volume is the biggest. 
2. The least volume of rides during the week is on Monday.
3. Members tend to have bigger volume of rides throughout the week, except on Saturday.
4. Casual's volume is the largest on Saturday which is ~56% compared to member which is only ~43%. 
5. Casual's volume increases during weekends, starting on friday with a ~7% increase. 

```{r plot 3}
ggplot(cyclistic_new, aes(weekday, fill = member_casual)) + geom_bar() + 
labs(x = "Weekday", title = "Plot 3: Weekday Distribution") + coord_flip() +
  scale_fill_brewer(palette = "Set2")
```

## Hour of the day

```{r hour of the day}
cyclistic_new %>% 
  group_by(start_hour) %>% 
  summarise(count = length(ride_id),
            '%' = (length(ride_id) / nrow(cyclistic_new)) * 100,
            'member_percent' = (sum(member_casual == "member") / length(ride_id)) * 100,
            'casual_percent' = (sum(member_casual == "casual") / length(ride_id)) * 100,
            'member_casual_perc_difer' = member_percent - casual_percent)
            
```
## Plot 4: Hour of the day distribution
From this data, there's a few key points worth noting:
1. The highest volume of rides is at 12pm afternoon. 
2. We have a bigger volume of rides during the afternoon. 
3. We have more members during the morning mainly starting at 5am-11am.
4. And we have casuals starting at 4pm-10pm.

```{r plot 4}
cyclistic_new %>% 
  ggplot(aes(start_hour, fill = member_casual)) + labs(x = "Hour of the day", title = "Plot 4: Hour of the day distribution") + geom_bar() + scale_fill_brewer(palette = "Set2")
```

## Plot 5: Hour of the day by weekday distribution
I will compare hour of the day throughout weekday using facet_wrap to see if there's 
any key points here:
1. We can tell that Mon, Tue, Wed, Thur, Fri data looks quite similar to each other.
2. And Sat and Sun data also look similar to each other.
3. There's an obvious difference between midweek data and weekend data. So let's
compare the midweek and weekend data together. 

```{r plot 5}
cyclistic_new %>% 
  ggplot(aes(start_hour, fill = member_casual)) + labs(x = "Hour of the day", title = "Plot 5: Hour of the day distribution divided by weekday") + geom_bar() + facet_wrap(~weekday) + scale_fill_brewer(palette = "Set2")
```

## Plot 6: Midweek and weekend distribution (need revision for description)
Since there's a clear difference between midweek and weekend. I decided to create 
another plot to better view the difference:
(the count of data points does not give as much meaning knowing each plot represents 
a different amount of days)
1. Flow- the weekend has a much smoother flow while the midweek has a steeper flow
of volume.
2. Midweek- there's an increase starting at 5am-12pm; then the data starts to decrease
starting at 1pm. Overall, there's more members than casuals throughout the day. 
3. Weekend- we have a much larger flow of casuals than members during the afternoon 
and evening. 

```{r plot 6}
cyclistic_new %>% 
  mutate(type_of_weekday = ifelse(weekday == '6 - Sat' | weekday == '7 - Sun',
                                  'weekend',
                                  'midweek')) %>% 
  ggplot(aes(start_hour, fill = member_casual))+ scale_fill_brewer(palette = "Set2") + labs(x = "Hour of the day", title = "Plot 6: Midweek and weekend distribution") + geom_bar() + facet_wrap(~type_of_weekday)
```

## Types of bike distribution
Let's compare to see which types of bikes have more preferences. 

```{r types of bike distribution}
cyclistic_new %>% 
  group_by(rideable_type) %>% 
  summarise(count = length(ride_id),
            '%' = (length(ride_id) / nrow(cyclistic_new)) * 100,
            'member_percent' = (sum(member_casual == "member") / length(ride_id)) * 100,
            'casual_percent' = (sum(member_casual == "casual") / length(ride_id)) * 100,
            'member_casual_perc_difer' = member_percent - casual_percent)
```

## Plot 7: Types of bike distribution
Based on the data, it's important to note a few key points here:
1. Both members and casuals tend to prefer classical and electric bikes more. However, this might be because we have more classical and electric bikes more. 
2. This data needs more information for better conclusion. 

```{r plot 7}
cyclistic_new %>% 
  ggplot(aes(rideable_type, fill = member_casual)) + labs(x= "Rideable type", title = 
                                                            "Plot 7: Type of bike distribution") + geom_bar() + coord_flip() + scale_fill_brewer(palette = "Set2")
```

## ride_time_m 
There a few problems that need to be addressed here:
1. The min is negative. How come the duration of the trip becomes negative? It
could be because of the stations' malfunction which return errors/ bad dates.
2. The max value also does not make sense. The mean here is 21.54 when the max is 
55944.15. 
3. Checking the start and end station doesn't appear to have a problem.

```{r ride_time_m}
summary(cyclistic_new$ride_time_m)
```

With the data below, we can see a few things here:
1. The difference between 0% and 100% is 56002 minutes.
2. The difference between 5% and 95% is ~52 minutes. 
3. Because of that, in the analysis of this variable we are going to use a subset of the dataset without outliners. The subset will contain 95% of the dataset.

```{r}
ventiles = quantile(cyclistic_new$ride_time_m, seq(0, 1, by=0.05))
ventiles
```

## Removed outliners

```{r removed outliners}
cyclistic_without_outliners <- cyclistic_new %>% 
  filter(ride_time_m > as.numeric(ventiles['5%'])) %>% 
  filter(ride_time_m < as.numeric(ventiles['95%']))
print(paste("Removed", nrow(cyclistic_new) - nrow(cyclistic_without_outliners), "rows as outliners"))
```

## Riding time for casual vs member distribution

```{r ride_time_m multivariable}
cyclistic_without_outliners %>% 
  group_by(member_casual) %>% 
  summarise(mean = mean(ride_time_m),
            'first_quarter' = as.numeric(quantile(ride_time_m, .25)),
            'median' = median(ride_time_m),
            'third_quarter' = as.numeric(quantile(ride_time_m, .75)),
            'IR' = third_quarter - first_quarter)
```

## Plot 8: Riding time for Casual x Member distribution
Based on the graph, we can see that casuals have more riding time than members, 
mean and IQR (InterQuartile Range) are aslo bigger for casuals. 

```{r plot 8}
cyclistic_without_outliners %>% 
  ggplot(aes(x= member_casual, y=ride_time_m, fill=member_casual)) + geom_boxplot() + labs(x="Member x Casual", y = "Riding time", title = "Plot 8: Riding time for Casual x Member distribution") + scale_fill_brewer(palette = "Set2")
```

## Plot 9: Riding time for day of the week distribution
Casuals- Riding times follow a more curve distribution, peaking on sundays and 
valleying on wednesday/thursday.
Members- Riding times keep unchanged starting from Monday to Thursday, we can see
the data starts to change starting from Friday.

```{r plot 9}
cyclistic_without_outliners %>% 
  ggplot(aes(x=weekday, y=ride_time_m, fill = member_casual)) + facet_wrap(~member_casual) +
geom_boxplot() + labs(x="Weekday", y="Riding time", title = "Plot 9: Riding time for the day of the week distribution") + coord_flip() + scale_fill_brewer(palette = "Set2")
```


## Plot 10: Riding time for rideable type distribution
I will not include this plot into the analysis since the information is not valid. 

```{r plot 10}
cyclistic_without_outliners %>% 
  ggplot(aes(x=rideable_type, y= ride_time_m, fill = member_casual)) +
  geom_boxplot() + coord_flip() + scale_fill_brewer(palette = "Set2") +
  facet_wrap(~member_casual) + labs(x= "Rideable type", y="Riding type",
                                    title = "Riding time for rideable type distribution")
```



